Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-10-24 ms/plos_latex.tex
still working on ms ...

The evolution of pathogen virulence has both 
theoretical and, potentially, practical
importance. In general, evolutionary theory suggests that
disease strains that can reproduce more --- where reproduction
is defined here as the amount of \emph{between-host} transmission, or
the number of new hosts infected --- will increase in prevalence.
Pathogens can increase their net reproduction rate either
by increasing their birth rate, the rate (per infected host) at which they
infect new hosts, or by decreasing their death rate, the rate
at which hosts recover or die from disease.
The \emph{trade-off theory} \cite{anderson_coevolution_1982,Ebert1999} --- postulates that
the between-host birth and death rate are both linked to the rate at which
the pathogen exploits host resources for within-host reproduction, 
and that pathogen
evolution will thus strike a balance between the
pathogen's rate of transmission
to new hosts and its rate of killing its host (or of provoking
the host's immune system to eliminate it).
Some biologists have criticized the tradeoff theory
\cite{EbertBull2003,alizon_adaptive_2015}, but others have
successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.
Fraser \etal\ have applied these ideas in a particularly
interesting way 
by showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in studies of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was both
heritable and covaried with the set-point viral load (i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection), which in turn predicted the probability of
transmission
\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately 100-200 years and then declining
slightly to a long-stable virulence level.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig1.pdf}
\caption{{\bf Baseline dynamics.}
Time series of mean population \Lspvl. (a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying exponential growth rates, $r$. (b) Effects of varying initial infectious density $I(0)$. (c) Effects of varying initial mean virulence $\alpha(0)$. The orange curve in panel (a) shows that our simplified model can produce similar trajectories to those Shirreff \etal's model \cite{shirreff_transmission_2011}. In order to compare the models, we match the initial exponential growth rate (see main text ``Simulation runs"). Furthermore, we experiment with wide ranges of initial values (panels b and c) to show that evolutionary dynamics are not overly sensitive to starting parameter values, which we hold constant in our simulations.}
\label{fig:panel3}
\end{figure}

@@

We construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$. Finally, the probabilities are normalized so that each row sums to 1:
\begin{equation}
M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\tsub{\alpha}{max} + d/2;i) - \Phi(\tsub{\alpha}{min} - d/2;i)},
\end{equation}
where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$. Transmission rate and disease induced mortality rates are discretized into a vector as well:
\begin{equation}
\begin{aligned}
\beta_i &= \beta(\alpha_i),\\
\lambda_i &= \frac{1}{D_P + D_A (\alpha_i) + D_D}.
\end{aligned}
\end{equation}

@@

We developed seven multi-strain evolutionary models covering a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics. Specifically, we focus on the effects of the assumptions of (1) instantaneous vs. non-instantaneous partnership formation, (2) zero vs. positive extra-partnership sexual contact and transmission and (3) homogeneous vs. heterogeneous levels of sexual activity on the evolution of mean \Lspvl.

@@

The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and simpler, implicit models, assume instantaneous partnership formation (``instswitch''). The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$. Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission.

The next two models represent extreme simplifications of sexual partnership dynamics.  One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}. It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics. However, to reflect the effect of partnership structure, it uses an adjusted transmission rate that is derived from an approximation of the basic reproduction number of a serial monogamy model with instantaneous pair formation \cite{hollingsworth_hiv1_2008}. The second model of this pair (``random'') is a simple random-mixing model.

Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero''). Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix for model details). We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}. While this model lacks some 
important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, we see it as a first step toward assessing the effects of epidemiological complexity. As even these models are pushing the limits of compartmental-based models, adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.

The base model (i.e. pairform+epc) for the first four models and the heterogeneous model is an extension of Champredon \etal's model \cite{champredon_hiv_2013}. Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$. Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate). Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing. Extra-couple and uncoupled transmission are modeled in the same way as Champredon \etal's model. All the details have been adapted to a multi-strain scenario. The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact). Model details are explained in the appendix.

@@

Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values. In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$). For the heterogeneity model, mean and squared coefficient of variation (CV) for number of non-cohabiting partners are sampled as well. We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.

@@

\begin{table}[h!]
\caption{Parameter ranges/values.  Values of $c$ and $\rho$ are doubled from those given by Champredon \etal; we keep track of individuals in the model, while they keep track of couples. Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
\centering
\begin{tabular}{c p{2in} c l}
\hline 
Notation & Description & Range/Value & Source\\
\hline % inserts single horizontal line
$\rho$ & Partnership formation rate & 1/10-2/5 per year & \cite{champredon_hiv_2013} \\
$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) per year & \cite{champredon_hiv_2013} \\
$c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
$c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) years & \cite{hollingsworth_hiv1_2008} \\
$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) years & \cite{hollingsworth_hiv1_2008} \\
$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 per year & \cite{shirreff_transmission_2011} \\
$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 copies per ml & \cite{shirreff_transmission_2011} \\
$\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
$\tsub{D}{max}$ & Duration of primary infection & 25.4 years & \cite{shirreff_transmission_2011} \\
$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 copies per ml & \cite{shirreff_transmission_2011} \\
$D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
$\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 7 & \cite{shirreff_transmission_2011}\\
$n$ & Number of strains & 21 (51*) & Assumption\\
$\mu$ & Mean number of non-cohabiting sexual partners & 0.103 - 1.206 & \cite{omori2015dynamics}\\
$\kappa$ & Squared coefficient of variation of number of non-cohabiting sexual partners & 0.01 - 100 & Assumption\\
\hline
\end{tabular}
\label{table:parmsTable}
\end{table}

The one completely new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix S1); we need to pick a reasonable range for it. Champredon \etal\ \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.  Given Champredon \etal's parameter range, the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5. Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$. Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- we consider the wider range appropriate, as little is not much known about the uncoupled transmission rate.

Two parameters, mean and the squared coefficient of variation (CV) of number of non-cohabiting partners, are sampled for heterogeneity in sexual activity. We decided to use squared CV instead of variance as variance measures is often unreliable \todo{CITE? You mentioned that variance is unreliable and we should use CV but we didn't talk about it very much...}. To allow for a wide range of uncertainty, range for the mean number of non-cohabiting partners was taken from unmarried men as that was the group with the largest variability \cite{omori2015dynamics}. However, we decided to use 0.01 - 100 as the sampling range for squared CV despite the wider CV range given by Omori \etal \cite{omori2015dynamics} (0 - 20.324). Although married female have largest variability for CV, the range of the mean is narrow and small, which might explain for such large variability. Assessing ranges of CV of other groups, we concluded that 0.01 - 100 is both reasonable and wide enough to cover uncertainties in parameters. For simplicity, we assume that the number of non-cohabiting partner follows a gamma distribution.

@@

One of the most difficult parts of model comparison is finding
parameter sets that are commensurate with many different model
structures. For the most part, our models are too complex to easily
derive analytical correspondences among them. Given a numerical
criterion, such as $r$ (initial exponential growth rate) or $\rzero$ 
(intrinsic reproductive number), we can adjust one or more
parameters by brute force to ensure that all of the models match
according to that criterion. While $\rzero$ is often considered
the most fundamental property of an epidemic, and might thus seem to
be a natural matching criterion, here we focus on matching the initial
growth rate $r$ for several reasons. First, our primary interest is in
the transient evolutionary dynamics of virulence, which are more
strongly affected by $r$ than $\rzero$. Second, $r$ is 
more directly observable in real epidemics; $r$ can be estimated by
fitting an exponential curve to the initial incidence or
prevalence curves \cite{ma_estimating_2014}, while $\rzero$
typically requires either (1) knowledge of \emph{all} epidemic
parameters or (2) calculations based on
$r$ and knowledge of the serial interval or generation interval of the
disease \cite{wallinga_how_2007}. Thus, we scale parameters so that
every run has the same initial exponential growth rate of incidence.

@@

We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
of mean \Lspvl\ of 3 and epidemic size of $10^{-4}$. After each run, initial exponential growth rate is calculated. Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that implied by Shirreff \etal's original models.

The calibration runs for each parameter set are slightly
simplified. We run each model for only 500 years
(full simulations are run for 4000 years). We use a 4/5 order 
Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
\cite{soetaert_solving_2010}) for all simulations.

@@

Knowing the minimum progression time, timing of the minimum progression time/peak virulence, and equilibrium progression time provide sufficient detail to identify the overall shape of the virulence trajectory.
In particular, knowing the timing of the peak virulence (how many years
into the epidemic the virulence peaks) can help epidemiologists
guess whether the virulence of an emerging pathogen is likely (1)
to have peaked early, possibly even before the pathogen is detected
spreading in the population, and decline over the remaining course
of the epidemic; (2) to increase, peak, and decline over the
foreseeable future; or (3) to increase very slowly, peaking only
in the far future. To the extent that our simplistic model for HIV
reflects reality, we would take the peak time of 150-300 years 
(Figure~\ref{fig:panel3}c) to mean that, in the absence of
treatment, the epidemic would probably still be increasing in virulence.

@@

Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases. Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b). Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence (\figurename~\ref{fig:panel3}c).

Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (\figurename~\ref{fig:virtraj}). Although the speed of virulence evolution varies, leading to wide variation in the minimum expected progression time (means ranging from approximately 6 to 12 years), virulence peaks in all models between 200 and 300 years.

@@

Our chosen summary statistics (peak time, minimum expected progression time, equilibrium expected
progression time, and relative progression time) all vary considerably across models
\ref{fig:unidist}.
We first consider the models of intermediate realism: implicit,
instantaneous-switching with and without extra-pair contact, and
pair formation without extra-pair contact. Some parameter
sets for these models lead to low equilibrium virulence ($\approx 18$ years
to progression);
these same sets lead to correspondingly low
peak virulence (16 years to progression) and early peak times (before 200 years).
At the opposite extreme, parameter sets that produce high equilibrium virulence (8 years to progression)
also produce late peaks ($> 200~\text{years}$) and
high peak virulence (4 years to progression).
The pair-formation without extra-pair contact and implicit models
occasionally have parameter sets that select for such low virulence across
the board that they never exceed their initial virulence, leading to a tail
of peak times near zero.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig3.pdf}
\caption{{\bf Univariate distributions of summary statistics.}
The distribution of equilibrium virulence for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.}
\label{fig:unidist}
\end{figure}

@@

\subsection*{Heteregenous model}

>>>>

The evolution of pathogen virulence has both 
theoretical and, potentially, practical
importance. In general, evolutionary theory suggests that
disease strains that can reproduce more --- where reproduction
is defined here as the amount of \emph{between-host} transmission, or
the number of new hosts infected --- will increase in prevalence.
Pathogens can increase their net reproduction rate either
by increasing their transmission rate, 
the rate (per infected host) at which they
infect new hosts, or by decreasing their clearance or disease-induced
mortality rate, the rate
at which hosts recover or die from disease.
The \emph{trade-off theory} \cite{alizon_virulence_2009} postulates that
the transmission and disease-induced mortality rate are both linked to the rate at which
the pathogen exploits host resources for within-host reproduction, 
and that pathogen
evolution will thus strike a balance between the
pathogen's rate of transmission
to new hosts and its rate of killing its host (or of provoking
the host's immune system to eliminate it).
Some biologists have criticized the tradeoff theory
\cite{EbertBull2003,alizon_adaptive_2015}, but others have
successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.
Fraser \etal\ have applied these ideas in a particularly
interesting way 
by showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in studies of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was both
heritable and covaried with the set-point viral load (i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection), which in turn predicted the probability of
transmission
\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately 100-200 years and then declining
slightly to a long-stable virulence level.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig1.pdf}
\caption{{\bf Baseline dynamics.}
Time series of mean population \Lspvl. (a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$. (b) Effects of varying initial infectious density $I(0)$. (c) Effects of varying initial mean virulence $\alpha(0)$. The $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's model \cite{shirreff_transmission_2011}, shows that our simplified model can produce similar virulence trajectories. Panels b and c illustrate the sensitivity of virulence trajectories to initial conditions $I(0)$ and $\alpha(0)$, which we hold constant in our simulations.}
\label{fig:panel3}
\end{figure}

@@

We construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$. Finally, the probabilities are normalized so that each row sums to 1:
\begin{equation}
M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\tsub{\alpha}{max} + d/2;i) - \Phi(\tsub{\alpha}{min} - d/2;i)},
\end{equation}
where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$. Transmission rate and disease induced mortality rates are discretized as well:
\begin{equation}
\begin{aligned}
\beta_i &= \beta(\alpha_i),\\
\lambda_i &= \frac{1}{D_P + D_A (\alpha_i) + D_D}.
\end{aligned}
\end{equation}

@@

We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics. Specifically, we focus on the effects of the assumptions of (1) instantaneous vs. non-instantaneous partnership formation; (2) zero vs. positive extra-partnership sexual contact and transmission; and (3) homogeneous vs. heterogeneous levels of sexual activity on the evolution of mean \Lspvl.

@@

The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and the simpler, implicit models used by Shirreff \emph{et al.} \cite{shirreff_transmission_2011}, assume instantaneous partnership formation (``instswitch''). The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$. Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission.

The fifth and sixth models represent extreme simplifications of sexual partnership dynamics.  One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}. It is actually a random mixing model that explicitly tracks only the total number of susceptible and infected individuals. However, to reflect the effect of partnership structure, it uses an adjusted transmission rate derived from an approximation of the basic reproduction number of a serial monogamy model with instantaneous pair formation \cite{hollingsworth_hiv1_2008}. The second model of this pair (``random'') is a simple random-mixing model.

Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero''). Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix). We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}. While this model lacks some 
important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity. As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises \todo{how many?} coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.

The pairform+epc and heterogeneous models use the basic epidemiological framework of Champredon \etal \cite{champredon_hiv_2013}. Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$. Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate). Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing. Extra-couple and uncoupled transmission are modeled in the same way as in Champredon \etal's model. All the details have been adapted to a multi-strain scenario, so that we track (for example) a matrix 
$II_{ij}$ that records the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$. 
The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact: see Appendix).

@@

Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values. In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$). For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well. We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), instead using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.

@@

\begin{table}[h!]
\caption{Parameter ranges/values.  Values of $c$ and $\rho$ are doubled from those given by Champredon \etal\ because we keep track of individuals in the model, while they keep track of couples. Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
\centering
\begin{tabular}{c p{2in} c l}
\hline 
Notation & Description & Range/Value & Source\\
\hline % inserts single horizontal line
$\rho$ & Partnership formation rate & 1/10-2/5 per year & \cite{champredon_hiv_2013} \\
$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) per year & \cite{champredon_hiv_2013} \\
$c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
$c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) years & \cite{hollingsworth_hiv1_2008} \\
$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) years & \cite{hollingsworth_hiv1_2008} \\
$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 per year & \cite{shirreff_transmission_2011} \\
$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 copies per ml & \cite{shirreff_transmission_2011} \\
$\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
$\tsub{D}{max}$ & Duration of primary infection & 25.4 years & \cite{shirreff_transmission_2011} \\
$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 copies per ml & \cite{shirreff_transmission_2011} \\
$D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
$\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 7 & \cite{shirreff_transmission_2011}\\
$n$ & Number of strains & 21 (51*) & Assumption\\
$\mu$ & Mean number of non-cohabiting sexual partners & 0.103 - 1.206 & \cite{omori2015dynamics}\\
$\kappa$ & Squared coefficient of variation of number of non-cohabiting sexual partners & 0.01 - 100 & Assumption\\
\hline
\end{tabular}
\label{table:parmsTable}
\end{table}

One parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix S1);
it appears neither in either Shirreff \etal\ nor Champredon \etal's models,  so we need to pick a reasonable range for it. Champredon \etal\ \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.  Given Champredon \etal's parameter range, the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5. Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$. Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- we consider the wider range appropriate, as little is not much known about the uncoupled transmission rate.

Two parameters, mean and the squared coefficient of variation (CV) of number of non-cohabiting partners, are sampled for heterogeneity in sexual activity.
%%  BMB: I don't think we need to go into detail about variance vs CV
To allow for a wide range of uncertainty, range for the mean number of non-cohabiting partners was taken from unmarried men, as that was the group with the largest variability \cite{omori2015dynamics}. 
Omori \etal \cite{omori2015dynamics} give a very wide
range for the coefficient of variation ($\approx$ 0 - 20, corresponding
to squared CV range of 0-400):
we narrowed this range for $\textrm{CV}^2$ to 0.01-100.
At the bottom end of the range, estimating that a group behaves
perfectly homogeneously ($CV=0$) is likely to be a sampling artifact;
at the upper end, the estimate is also likely to be noisy because
of the low mean value among married females (who have the largest
range of CV). 
We assume that the number of non-cohabiting partners follows a Gamma distribution.

@@

One of the most difficult parts of model comparison is finding
parameter sets that are commensurate with many different model
structures. For the most part, our models are too complex to easily
derive analytical correspondences among them. Given a numerical
criterion, such as $r$ (initial exponential growth rate) or $\rzero$ 
(intrinsic reproductive number), we can adjust one or more
parameters by brute force to ensure that all of the models match
according to that criterion. While $\rzero$ is often considered
the most fundamental property of an epidemic, and might thus seem to
be a natural matching criterion, here we focus on matching the initial
growth rate $r$ for several reasons. First, our primary interest is in
the transient evolutionary dynamics of virulence, which are more
strongly affected by $r$ than $\rzero$. Second, $r$ is 
more directly observable in real epidemics; $r$ can be estimated by
fitting an exponential curve to the initial incidence or
prevalence curves \cite{ma_estimating_2014}, while $\rzero$
typically requires either (1) knowledge of \emph{all} epidemic
parameters or (2) calculations based on
$r$ and knowledge of the serial interval or generation interval of the
disease \cite{wallinga_how_2007}. Thus, we scale parameters so that
every run has the same initial exponential growth rate in 
disease incidence.

@@

We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
of mean \Lspvl\ of 3 and epidemic size of $10^{-4}$. After each run, initial exponential growth rate is calculated. Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, a value that approximates the growth rates of Shirreff \etal's original models.
For calibration purposes, we run each model for only 500 years
(full simulations are run for 4000 years), which is always long
enough to capture the exponential growth phase of the model. 
We use a 4/5 order 
Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
\cite{soetaert_solving_2010}) for all simulations.

@@

Knowing the minimum progression time, timing of the minimum progression time/peak virulence, and equilibrium progression time provide sufficient detail to identify the overall shape of the virulence trajectory.
In particular, knowing the timing of the peak virulence (how many years
into the epidemic the virulence peaks) can help epidemiologists
guess whether the virulence of an emerging pathogen is likely (1)
to have peaked early, possibly even before the pathogen is detected
spreading in the population, and decline over the remaining course
of the epidemic; (2) to increase, peak, and decline over the
foreseeable future; or (3) to increase very slowly, peaking only
in the far future. To the extent that our simplistic model for HIV
reflects reality, we would take the peak time of 150-300 years 
(\figurename~\ref{fig:panel3}c) to mean that, in the absence of
treatment, the epidemic would probably still be increasing in virulence.

@@

Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases. Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b). Decreasing the initial virulence
also leads to progressively later, larger peaks in virulence (\figurename~\ref{fig:panel3}c).

Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories, which we quantify in terms of the expected time of progression
to AIDS (\figurename~\ref{fig:virtraj}: lower progression time corresponds
to higher virulence). Although the speed of virulence evolution varies, leading to wide variation in the minimum expected progression time (means ranging from approximately 6 to 12 years), virulence peaks in all models between 200 and 300 years.

@@

Our chosen summary statistics (peak time, minimum expected progression time, equilibrium expected
progression time, and relative progression time) all vary considerably across models
(\figurename~\ref{fig:unidist}).
We first consider the models of intermediate realism: implicit,
instantaneous-switching with and without extra-pair contact, and
pair formation without extra-pair contact. Some parameter
sets for these models lead to low equilibrium virulence ($\approx 18$ years
to progression);
these same sets lead to correspondingly low
peak virulence (16 years to progression) and early peak times (before 200 years: \figurename~\ref{fig:pairplot}).
At the opposite extreme, parameter sets that produce high equilibrium virulence (8 years to progression)
also produce late peaks ($> 200~\text{years}$) and
high peak virulence (4 years to progression).
The pair-formation without extra-pair contact and implicit models
occasionally have parameter sets that select for such low virulence across
the board that they never exceed their initial virulence, leading to a tail
of peak times near zero.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig3.pdf}
\caption{{\bf Univariate distributions of summary statistics.}
The distribution of equilibrium expected progression
time (lower left panel) for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.}
\label{fig:unidist}
\end{figure}

@@

\subsection*{Heterogenous model}
